Kindly Bent to Free Us

### 2018-11-15 kindly/infer/infer.tex
Small fixes.

We now formulate type inference for the \lang language. Our inference technique is based on the HM(X) framework~\citep{DBLP:journals/tapos/OderskySW99} which
presents how to implement Hindley-Milner type inference in a language
with qualified types. We first present various preliminaries definitions,
we then show our extension of HM(X) that support a kind system and affine
types.

@@

We consider $\mathcal C$ the set of constrains.
The grammar of constrains, presented
in \cref{grammar:constraint}, follows the traditional HM(X) formulation
with conjunctions,projections and type inequalities. The only new
element specific to our approach are kind equalities.
Entailment is noted $\entail{C}{D}$, where $D$ is a consequence of the
constraints $C$.
The base entailment rules are shown in \cref{rules:entail}, we consider
the completion of that system to form a cylindric constraint system.
\TODO{Give the properties}

We say that $C$ and $D$ are equivalent, noted $C \equivC D$,
when $\entail{C}{D}$ and $\entail{D}{C}$.

@@

We note $\mathcal S$ the set of solved forms
which can be used inside type and kind schemes.
We define $\mathcal S$ as $\mathcal C$ quotiented by the relation $\equivC$.
%
We consider the existence of a function $\normalize$ which takes
a constraint in $\mathcal C$ and a unifier $\psi$ and returns a constraint
in solved form $C' \in \mathcal S$,
and an updated unifier. We detail the implementation
of the normalization function in \cref{sec:normalize}

@@

\newcommand\lub\bigvee
\newcommand\glb\bigwedge

We now place ourselves in a more general context where kinds are either variables
or constants belonging to a total bounded lattice $(\mathcal L, \leq)$ ie.,
a lattice which admits a total order and upper and lower bounds ($l^\top$ and $l^\bot$).
We note lattice elements $l$ and $\glb_i l_i$ (resp. $\lub_i l_i$)
the greatest lower bound (resp. least upper bound) in $\mathcal L$.
%
Let $\mathcal C_{\mathcal L}$ the set of constraints in this lattice. We consider the
following base entailment rules in this system:
\begin{center}
  \begin{mathpar}
    \inferrule{l \leq l'}{\entail{}{\Cleq{l}{l'}}}
    \and
    \inferrule{}{\entail{}{\Cleq{k}{l^\top}}}
    \and
    \inferrule{}{\entail{}{\Cleq{l^\bot}{k}}}
  \end{mathpar}
\end{center}

The normalization procedure is noted $\normalize{C_0}{\unif_0}$ where
$C_0\in \mathcal C_{\mathcal L}$ is a set of constraints and $\unif_0$ is a substitution.
It returns a constraint $C \in \mathcal S$ in
solved form and a unifier $\unif$.
The main idea of the algorithm is to first remove all the type equalities
by using regular Herbrand unification. After that, we only have
a set of inequalities among kinds, which we can consider as a relation.
We can then saturate the relation,
unify all kinds that in the same equivalence classes to obtain
a most general unifier on kind variables,
remove all existentially quantified variables and
then minimize back the relation.

@@

\begin{lemma}[Normal form]
  \label{lemma:normalform}
  Given a constraint $D\in\mathcal C$, a substitution $\phi$ and
  $(C,\unif) = \normalize{D}{\phi}$,
  then $\phi\leq\unif$,
  $C \equivC \unif D$ and
  $\unif C = C$.
\end{lemma}
\begin{proof}
  % For simplicity, we assume that any substitution has been already applied
  % to $D$ and that $\phi = id$.
  Let us partition $\phi$ into a part which affects type variables,
  $\phi_\tau$, and a part which affects kind variables, $\phi_k$.

@@

\begin{theorem}
  The $\operatorname{normalize}$ procedure is complete and computes
  unique principal normal forms for our constraint system.
\end{theorem}

>>>>

We now formulate type inference for the \lang language. Type inference is
based on the HM(X) framework~\citep{DBLP:journals/tapos/OderskySW99} which
presents how to implement Hindley-Milner type inference in a language
with qualified types. We first present various preliminaries definitions,
we then show our extension of HM(X) that support a kind system and affine
types.

@@

\newcommand\A{\mathcal A}
\newcommand\SC{\mathcal S}

Let us note $\A$ our constraint system. The grammar of constraints is
given in  \cref{grammar:constraint}.
$\A$ is defined as the smallest cylindric term constraint system that
satisfies the axiom shown in \cref{rules:entail}.
We follows the traditional HM(X) formulation
with conjunctions, projections and type inequalities.
The new element specific to our approach are kind inequalities.
Entailment is noted $\entail{C}{D}$, where $D$ is a consequence of the
constraints $C$.
We say that $C$ and $D$ are equivalent, noted $C \equivC D$,
when $\entail{C}{D}$ and $\entail{D}{C}$.
\TODO{Give the cylindric properties ?}

@@

We note $\SC$ the set of solved forms
which can be used inside type and kind schemes.
We define $\SC$ as $\A$ quotiented by the relation $\equivC$.
%
We consider the existence of a function $\operatorname{normalize}$ which takes
a constraint in $\A$ and a substitution $\psi$ and returns a constraint
in solved form $C' \in \SC$,
and an updated substitution. We detail the implementation
of the normalization function in \cref{sec:normalize}

@@

\newcommand\lub\bigvee
\newcommand\glb\bigwedge
\newcommand\CL{{\mathcal C_{\mathcal L}}}

We now place ourselves in a more general context where kinds are either variables
or constants belonging to a total bounded lattice $(\mathcal L, \leq_{\mathcal L})$ ie.,
a lattice which admits a total order and upper and lower bounds ($l^\top$ and $l^\bot$).
We note lattice elements $l$ and $\glb_i l_i$ (resp. $\lub_i l_i$)
the greatest lower bound (resp. least upper bound) in $\mathcal L$.
%
Let $\CL$ the set of constraints in this lattice. We consider the
following base entailment rules in this system:
\begin{center}
  \begin{mathpar}
    \inferrule{l \leq_{\mathcal L} l'}{\entail{}{\Cleq{l}{l'}}}
    \and
    \inferrule{}{\entail{}{\Cleq{k}{l^\top}}}
    \and
    \inferrule{}{\entail{}{\Cleq{l^\bot}{k}}}
  \end{mathpar}
\end{center}

The normalization procedure is noted $\normalize{C_0}{\unif_0}$ where
$C_0\in \CL$ is a set of constraints and $\unif_0$ is a substitution.
It returns a constraint $C \in \mathcal S$ in
solved form and a unifier $\unif$.
The main idea of the algorithm is to first remove all the type equalities
by using regular Herbrand unification. After that, we only have
a set of inequalities among kinds, which we can consider as a relation.
We can then saturate the relation,
unify all kinds that in the same equivalence classes to obtain
a most general unifier on kind variables,
remove all existentially quantified variables and
then minimize back the relation.

@@

\subsubsection{Soundness and Principality}

\begin{lemma}[Normal form]
  \label{lemma:normalform}
  Given a constraint $D\in\CL$, a substitution $\phi$ and
  $(C,\unif) = \normalize{D}{\phi}$,
  then $\phi\leq\unif$,
  $C \equivC \unif D$ and
  $\unif C = C$.
\end{lemma}
\begin{proof}
  % For simplicity, we assume that any substitution has been already applied
  % to $D$ and that $\phi = id$.
  Let us partition $\phi$ into a part which affects type variables,
  $\phi_\tau$, and a part which affects kind variables, $\phi_k$.

@@

\begin{theorem}
  $\CL$ has the principal constraint property and
  $\operatorname{normalize}$ computes principal normal forms for $\CL$.
\end{theorem}

\begin{lemma}
  For all $C\in\mathcal S$, $\entail{C}{x = x}$ implies
  $\entail{}{x = x}$.
\end{lemma}
\begin{proof}
  By definition of $\operatorname{normalize}$, We have $C = \Cleq{k}{k'}_i$
  such that the underlying relation has no cycles.
  Thus, we can not deduce neither kind nor type equalities from $C$.
\end{proof}

\begin{lemma}
  $\CL$ is regular, ie, for $x, x'$ two types or kinds,
  $\entail{}{\Ceq{x}{x'}}$ implies
  $\fv{x} = \fv{x'}$
\end{lemma}
\begin{proof}
  The only equalities possibles are between variables (via symmetry) or
  between constants.
\end{proof}

\begin{theorem}
  Type inference in $HM(\CL)$ is sound, principal and complete.
\end{theorem}
